A sustainable approach to tourist signage on heritage trails\footnote{This research has been supported by the Ministry of University and Research under the "Underlandscape" project (2020428LS8)}

### 2023-11-08 paper.tex
Post update 8/11 (serale)

Its use is especially appropriate when the cultural heritage extends across a scarcely accessible land, preserving and accessing sites located on private property or in hazardous areas can be challenging. In these cases, QR tags could improve the visitors' experience by providing, besides textual data, the Web URLs for videos, aerial shots, and photos. Multimedia resources allow visitors to immerse themselves in the area's history and gain an experiential understanding, even if they don't have access to a specific spot. The resulting widespread knowledge can be a beneficial tourist-economic engine that encourages investments in cultural heritage policies, thus achieving the goal of protecting and conserving abandoned sites.

@@

For a benchmarking point of view, the provincial touristic context of \emph{Lucca} is mentioned in this study with the purpose of highlighting the percentage weight of \emph{Bagni di Lucca} tourism flows within the whole tourism dynamics of the Province of \emph{Lucca}. 
As concerns tourist indexes shown in Chart 3, the touristic density counts 1866.96 people per square kilometer (1773 $km^2$), with an average stay of 3.36 nights per tourist, showing a medium-length permanence in the provincial territory; this data is in line with \emph{Bagni di Lucca} average stay (2.98 nights). 
The provincial area of \emph{Lucca} shows an accommodation density of 19.30 on 10.14 of \emph{Bagni di Lucca}, while the coexistence index of \emph{Lucca} (with 69.87 foreign tourists on 100 Italian tourists) is lower than \emph{Bagni di Lucca}’s coexistence index (with 103.74 foreign tourists in 100 Italian tourists). 
Definitely, it can be claimed that \emph{Bagni di Lucca} is well positioned in the whole touristic context of \emph{Lucca} since it represents one of the less populated municipalities in the territory with interesting tourist indexes. In fact, \emph{Bagni di Lucca} boasts its historical thermal tourism tradition together with a well-equipped environment for sport and adventure tourism, on a surface of 164.70 $km^2$, that is one of the biggest municipality areas in the province of \emph{Lucca}.

Several promotional websites advertise Bagni di Lucca attractions such as these two destination websites: www.turismobagnidilucca.com and www.valdilima. org; the sitography closing this paper contains a long yet incomplete list that confirms such a statement.

From the point of view of our research, the context of Bagni di Lucca helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as: anthropic, socio-cultural, and historical features, as well as touristic services). Indeed, Val di Lima can promote a very identifying destination brand straddling the town of Lucca and the Garfagnana mountain area. The Municipality of Bagni di Lucca and the tourist area of the Val di Lima are the context for the small and fascinating village of Casoli, which is one of the 31 villages belonging to the Municipality. It is 7.22 km far from Bagni di Lucca, at an altitude of 500 meters.

Casoli village boasts a heterogeneous range of tourist attractions, including both soft and hard outdoor experiential activities, such as: canyoning, rafting, and trekking experiences in the Canyon park. Additionally, it features Romanesque churches and huts attracting visitors looking for various experiences such as thermal, environmental, outdoor, sport and cultural activities. Throughout the year, local stakeholders offer a variety of services, often incorporating sustainable tourism practices. After a detailed analysis of the tourist and social context of Bagni di Lucca, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of touristic experiences \cite{loz09} dating back to the ‘90s, when some associations \citeWeb{agu,r2o,ronda,garfraft,valdilima,canyonpark,experience,ebikea,ebikel,bdlt2,bdl} began organizing mainly trekking excursions and rafting experiences .

As part of Bagni di Lucca's ecosystem, with its focus on tourism, Casoli is a destination for slow travel. It is important to ensure that any development actions taken are sustainable, in order to preserve its unique social fabric. This means paying attention to conservation of the environment and cultural heritage, as well as preserving intergenerational continuity and economic equity \cite{kuh10a}.

%Summing up, the quantitative data give evidence of a vibrant and attractive tourist reality constantly attracting tourists during the last decade. Thanks to such tourist flows \emph{Bagni di Lucca} is considered one of the main tourist areas of the \emph{Val di Lima} and the Serchio Valley. Several promotional websites advertise \emph{Bagni di Lucca} attractions\footnote{Among local websites, these two destination websites are: \url{www.turismobagnidilucca.com} and \url{www.valdilima.org}}: the sitography closing this paper contains a long yet incomplete list that confirms such a statement, and Table \ref{tab:experience} exposes the promotional touristic system offering varied experiences in the \emph{Val di Lima} area. The variety of outdoor touristic services dates back to the ‘90s when some associations began organizing mainly trekking excursions and rafting experiences.
%
%From the point of view of our research, the context of the \emph{Bagni di Lucca} helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as: anthropic, socio-cultural, and historical features, as well touristic services).
%
%After a detailed analysis of the touristic and social context of \emph{Bagni di Lucca}, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of experiences, such as thermal, environmental, outdoor, sport and cultural activities  \cite{loz09}.
%
%Indeed, \emph{Val di Lima} can promote a very identifying destination brand straddling the town of Lucca and the \textit{Garfagnana} mountain area. 
%
%%The strategic position of \emph{Bagni di Lucca} is one of the strengths of its tourism context, together with the presence of such attractive elements like Orrido di Botri, with The Canyon Park \footnote{\url{ https://www.canyonpark.it/}}, and other service companies and associations for experiential tourism.
%
%\begin{table}
%    \centering
%    \begin{tabular}{|p{70mm}||p{45mm}|} \hline 
%         {\bf Touristic promotion }&  {\bf Kind of service}\\
%         \hline \hline 
%         Aguarajà River Experience \newline \url{https://aguaraja.it/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\
%         \hline 
%         Rafting H2O \newline \url{https://www.raftingh2o.com/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\
%         \hline 
%         RocKonda \newline \url{https://www.rockonda.it/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\ \hline 
%         A.S.D. Garfagnana Rafting \newline \url{https://www.garfagnanarafting.com/}& Soft rafting, Rafting, Canyoning, aqua-trekking, kayak\\ 
%        \hline 
%         Val di Lima Off-Road \newline \url{https://www.valdilimaoffroad.com} & Quad, enduro\\ \hline 
%         Canyon Park \newline \url{https://www.canyonpark.it/canyon-sup}& Canyoning, aqua-trekking, stand-up padding, mental care and yoga, eco laboratory\\
%         \hline 
%         Pro Loco Bagni di Lucca \newline  \url{https://www.turismobagnidilucca.com/} \newline \url{https://www.valdilima.org/}& Sportfishing, mental care, trekking, survival, paragliding, horse riding, diving\\
%         \hline 
%         eBike Adventure Tour \newline \url{http://www.ebikeadventuretour.it/}& Ebike, Mountain bike\\ \hline 
%         Val di Lima eBike \newline  \url{https://valdilimaebike.it/} & Mountain e-bike, Quad, Enduro, Paintball\\
%         \hline 
%        Terme Bagni di Lucca \newline \url{https://www.termebagnidilucca.it/} &Spa, beauty and wellness experience\\
%        \hline
%    \end{tabular}
%    \caption{\textbf{Tourism experiences’ offer in \emph{Val di Lima} area \tablefootnote{The list might not include all private stakeholders of the territory, since some touristic operators might have an Internet presence and others might operate as public-private players. The companies and associations in the list have been individually contacted but some did not respond to our survey on touristic information.}}}
%    \label{tab:experience}
%\end{table}
%
%The Municipality of \emph{Bagni di Lucca} and the tourist area of the \emph{Val di Lima} are the context for the small village of \emph{Casoli}, which is one of the 31 villages belonging to the Municipality. It is 7,22 km far from \emph{Bagni di Lucca}, at an altitude of 500 meters of altitude.
%
%\emph{Casoli} is a fascinating village settled on the green and rolling hill landscapes crossed by the river \emph{Lima}, in the \emph{Media Valle del Serchio}. Its territory features multiple kinds of touristic sights: from soft and hard outdoor experiential attractions (specifically the Canyon park activities like canyoning, rafting, and trekking experiences)) to Romanesque churches’ remains and huts. Definitely, despite its variety of tourist attractions, \emph{Casoli} exhibits unique features in its comprehensive tourism.
%
%%There aren’t touristic information at the administrative level of \emph{Casoli} , so that the tourism context analysis concerns the Municipality area of \emph{Bagni di Lucca}, as the first territorial context providing tourist data registered by ISTAT.
%
%Given the context and the local vocation, we conclude that \emph{Casoli} is a small destination for tourists practicing heterogeneous touristic circuits in individual and combined ways. According to qualitative and quantitative analysis \cite{loz09} on the touristic and social context of \emph{Bagni di Lucca}, as the municipal area encompassing \emph{Casoli} village, the mentioned area boasts a good touristic appeal towards Italian and foreign tourists searching for various kinds of experiences, such as thermal, environmental, outdoor, sport and cultural activities. Local stakeholders operating in the tourism system offer a variety of services all year round. Whenever possible the offer is enriched with original touristic approaches paying attention to sustainability. As part of the \emph{Bagni di Lucca}’s ecosystem with its touristic vocation, \emph{Casoli} is defined as a well-known destination. Considering that \emph{Casoli} is a destination for slow tourism, ensuring the sustainability of development actions is crucial to maintaining its unique social fabric. The latter aspect suggests paying attention to environmental conservation, cultural heritage conservation, intergenerational continuity, and economic equity \cite{kuh10}.

@@

As described by \cite{dre22} Dianne Dredge and in light of the above analysis, private and public entities can contribute to sustainable tourism  by developing the following actions:

%In this perspective, policy-makers play a key role in the planning of sustainable measures and in the monitoring of impacts; in particular, in the context of tourism, sustainability is a topic discussed in its many aspects, including resident ecosystem resilience approaches and regeneration actions by policy-makers and/or private tour operators \cite{dre22}.

@@

Regarding residents, we had the chance to meet a few people living in \emph{Casoli}, which did not provide useful information; local entrepreneurs we met informed us about the trekking tourism flows passing through \emph{Casoli}  and the Piella Cave path, which usually starts from the Canyon Park experience.

@@

\subsubsection{Sites and trail’s state of the art \ref{sec:soa}}

@@

\begin{figure}
\includegraphics[width =\linewidth]{figure/combo_gen.jpeg}
\caption{Generation of the QR tag and UMap visualization from UMap content. The uMap editor (bottom left) provides a simplified access to the GeoJSON description. The same interface allows to specify the template controlling the Web visualization of the map (right side). An ad-hoc application referenced in the paper allows to transform the uMap file downloaded from the server into a batch of printable QR tags (top left). The tag is highlighted in red in the picture on the Web display}
\end{figure}

@@

Editing the map downloaded from the GPS app to fill in the {\em properties} does not require technical skills, since the tools integrated into the UMap user interface already fit the purpose (see figure \ref{fig:umap-editing}). The resulting \emph{GeoJSON} object describing a tagged feature is shown in figure \ref{fig:geojson-sample}.

The master document is rendered by UMap on the page linked to the tags (as in figure \ref{fig:linkmap}).

@@

To avoid the over-tourism pitfall, we need to take into account the other three pillars: social, environmental, and economic. To this end we analyze the social fabric that will host our initiative, the environment where signage will be deployed, and the economic framework that will support its implementation. From this study we derive the requirements for our signage system: low environmental impact, open to participation of the residents, easy to use from the visitors, and economically sustainable with a minimal touristic flow.

The approach we applied to reach such conclusions is summarized below: it is conceptual and therefore applicable to a wide range of use cases, and merely consists in designing the system with sustainability in mind from the early stages.

Our design started by defining a profile for the area according to the \emph{three pillars}. The environmental aspect is characterized by natural formations and human artifacts tracing back to the Paleolithic Era. As for tourism-related economics, a qualitative and quantitative analysis \cite{loz09} reveals unique features that need to be capitalized on to enhance the tourism potential. Given the touristic trends and the local vocation, we conclude that Casoli is a small destination for visitors practicing heterogeneous touristic circuits in individual and combined ways. Cultural sustainability mandates the involvement of local stakeholders in planning new heritage itineraries and supporting initiatives, including signage.

An outcome of the analysis of the social fabric is that, in order to preserve the native social and economic framework, mass tourism is not a candidate target, while a motivated, non-casual visitor is preferable. Regarding the potential reasons for a visit, we found a cave inhabited since pre-history but reachable with some effort, a village with historical buildings, and nearby architecture dating back to the Middle Ages. Such targets offer a variety of access ways, from those requiring performance to the relaxing ones. For this reason we opted for not suggesting a geo-itineray, allowing the user to follow personal inclinations. In this way we expect, justified by the results of \cite{sol18a}, that the educational impact of the signage is maximized.

@@

We show how a balanced use of technology may help in the task, with a process that is necessarily multidisciplinary, joining skills coming from the diverse domains: sociology, economics, humanities, and engineering. The paper demonstrates how their interplay returns ad-hoc tools targeting sustainability and efficacy.

@@

\nociteWeb{*}
\bibliographystyleWeb{abbrv}
\bibliographyWeb{biblio/sitography.bib}

% If authors have biography, please use the format below
%\section*{Short Biography of Authors}
%\bio
%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author1.pdf}}}
%{\textbf{Firstname Lastname} Biography of first author}
%
%\bio
%{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/author2.jpg}}}
%{\textbf{Firstname Lastname} Biography of second author}

% For the MDPI journals use author-date citation, please follow the formatting guidelines on http://www.mdpi.com/authors/references
% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p. 328; \citeyear{ref-journal-3b}, p.475). This produces: Wong (1999, p. 328; 2000, p. 475)

>>>>

Its use is especially appropriate when the cultural heritage extends across a scarcely accessible land, preserving and accessing sites located on private property or in hazardous areas can be challenging. In these cases, QR tags could improve the visitors’ experience by providing, besides textual data, the Web URLs for videos, aerial shots, and photos, even if they don’t have access to a specific spot. Multimedia resources allow visitors to immerse themselves in the area's history and gain an experiential understanding, even if they don't have access to a specific spot. The resulting widespread knowledge can be a beneficial tourist-economic engine that encourages investments in cultural heritage policies, thus achieving the goal of protecting and conserving abandoned sites.

@@

For a benchmarking point of view, the provincial touristic context of \emph{Lucca} is mentioned in this study with the purpose of highlighting the percentage weight of \emph{Bagni di Lucca} tourism flows within the whole tourism dynamics of the Province of \emph{Lucca}. 
As concerns tourist indexes shown in Table \ref{tab:tourism}, the touristic density counts 1866.96 people per square kilometer (1773 $km^2$), with an average stay of 3.36 nights per tourist, showing a medium-length permanence in the provincial territory; this data is in line with \emph{Bagni di Lucca} average stay (2.98 nights). 
The provincial area of \emph{Lucca} shows an accommodation density of 19.30 on 10.14 of \emph{Bagni di Lucca}, while the coexistence index of \emph{Lucca} (with 69.87 foreign tourists on 100 Italian tourists) is lower than \emph{Bagni di Lucca}’s coexistence index (with 103.74 foreign tourists in 100 Italian tourists). 
Definitely, it can be claimed that \emph{Bagni di Lucca} is well positioned in the whole touristic context of \emph{Lucca} since it represents one of the less populated municipalities in the territory with interesting tourist indexes. In fact, \emph{Bagni di Lucca} boasts its historical thermal tourism tradition together with a well-equipped environment for sport and adventure tourism, on a surface of 164.70 $km^2$, that is one of the biggest municipality areas in the province of \emph{Lucca}.

Several promotional websites, like \citeWeb{tbdl} and \citeWeb{valdilima} advertise \emph{Bagni di Lucca} attractions. The sitography closing this paper contains a longer yet incomplete list of touristic initiatives  \citeWeb{agu,r2o,ronda,garfraft,valdilima,canyonpark,experience,ebikea,ebikel,bdlt2,bdl}.

From the point of view of our research, the context of \emph{Bagni di Lucca} helps to understand the complexity and variety of mixed tourism clusters (a sort of "touristic ecosystems" with common identitary elements such as: anthropic, socio-cultural, and historical features, as well as touristic services). Indeed, \emph{Val di Lima} can promote a very identifying destination brand straddling the town of \emph{Lucca} and the \emph{Garfagnana} mountain area. The Municipality of \emph{Bagni di Lucca} and the tourist area of the \emph{Val di Lima} are the context for the small and fascinating village of \emph{Casoli}, which is one of the 31 villages belonging to the Municipality. It is 7.22 km far from \emph{Bagni di Lucca}, at an altitude of 500 meters.

\emph{Casoli} village boasts a heterogeneous range of tourist attractions, including both soft and hard outdoor experiential activities, such as: canyoning, rafting, and trekking experiences in the Canyon Park. Additionally, it features Romanesque churches and huts attracting visitors looking for various experiences such as thermal, environmental, outdoor, sport and cultural activities. Throughout the year, local stakeholders offer a variety of services, often incorporating sustainable tourism practices. After a detailed analysis of the tourist and social context of \emph{Bagni di Lucca}, it can be claimed that the area boasts a good tourist appeal towards Italian and foreign tourists searching for various kinds of touristic experiences \cite{loz09} dating back to the ‘90s, when some associations began organizing mainly trekking excursions and rafting experience .

As part of \emph{Bagni di Lucca}'s ecosystem, with its focus on tourism, \emph{Casoli} is a destination for slow travel. It is important to ensure that any development actions taken are sustainable, in order to preserve its unique social fabric. This means paying attention to conservation of the environment and cultural heritage, as well as preserving intergenerational continuity and economic equity \cite{kuh10a}.

@@

Based on the tourism trend of \emph{Casoli} and its overall tourism, and in alignment with the research goals focused on sustainable tourism, it is crucial to consider the role of local policy-makers. In fact, they play a vital role in planning sustainable measures and monitoring their impact.

In these terms, as described by [50] Dianne Dredge, and in light of the above analysis, private and public entities can contribute to sustainable tourism by developing the following actions:

@@

Regarding residents, we had the chance to meet a few people living in \emph{Casoli}, which did not provide useful information; local entrepreneurs we met informed us about the trekking tourism flows passing through \emph{Casoli} and the \emph{Piella} cave path, which usually starts from the Canyon Park experience.

@@

\subsubsection{Sites and trail’s state of the art \label{sec:soa}}

@@

\begin{figure}
\includegraphics[width =\linewidth]{figure/combo_gen.jpeg}
\caption{Generation of the QR tag and UMap visualization from UMap content. The uMap editor (bottom left) provides a simplified access to the GeoJSON description. The same interface allows to specify the template controlling the Web visualization of the map (right side). An ad-hoc application referenced in the paper allows to transform the uMap file downloaded from the server into a batch of printable QR tags (top left). The tag is highlighted in red in the picture on the Web display \label{fig:editing}}
\end{figure}

@@

Editing the map downloaded from the GPS app to fill in the {\em properties} does not require technical skills, since the tools integrated into the UMap user interface already fit the purpose (see the bottom left box in figure \ref{fig:editing}). The resulting \emph{GeoJSON} object describing a tagged feature is shown in figure \ref{fig:geojson-sample}.

The master document is rendered by UMap on the page linked to the tags (as shown in figure \ref{fig:editing}).

@@

To avoid the over-tourism pitfall, we need to take into account the other three pillars: social, environmental, and economic. To this end we analyze the social fabric that will host our initiative, the environment where signage will be deployed, and the economic framework that will support its implementation. From this study we derive the requirements for our signage system: low environmental impact, open to participation of the residents, responsive to the visitor's information needs, and economically sustainable with a minimal touristic flow.

The approach we applied to reach such conclusions is summarized below: it is conceptual and therefore applicable to a wide range of use cases, and merely consists in designing the system with sustainability in mind from the early stages. The simplicity of the approach description must not mislead: its implementation requires collaboration among experts in distinct disciplines converging to the final result.

Our design started by defining a profile for the area according to the \emph{three pillars}. The environmental aspect is characterised by natural formations and special habitats hosting a vast biodiversity, placed under special protection according to European legislation.

As for tourism-related economics, a qualitative and quantitative analysis \cite{loz09} reveals unique features that need to be capitalized on to enhance the tourism potential. In this light, community-involved tourism
represents a virtuous approach for destination management in inland areas that need a renewed sustainability-oriented design. Therefore, public and private stakeholders are crucial for managing spatial planning and addressing local issues through investments, conservation, promotion and monitoring actions.

In this perspective, our research is oriented to stimulate local policymakers and tourism professionals towards a better environmental and cultural sustainability approach by taking advantage of our study results to realize new potential heritage trails in Casoli and its vicinity.

%As for tourism-related economics, a qualitative and quantitative analysis reveals unique features that need to be capitalized on to enhance the tourism potential.

An outcome of the analysis of the social fabric is that, in order to preserve the native social and economic framework, mass tourism is not a candidate target, while a motivated, non-casual visitor is preferable. Regarding the potential reasons for a visit, we found a cave inhabited since pre-history but reachable with some effort, a village with historical buildings, and nearby architecture dating back to the Middle Ages. Such targets offer a variety of access ways, from those requiring performance to the relaxing ones. For this reason we opted for not suggesting a geo-itineray, allowing the user cherry-picking locations based on personal interests. In this way we expect, legitimated by the results of \cite{sol18a}, that the educational impact of the signage is maximized. Furthermore, our choice aims at a conciseness and communicative effectiveness of the tourist information, by inserting within the QR code a text of an average of 460 characters that facilitates readability and immediately provides the information without further steps to web pages, which may slow down the visit or discourage the less motivated visitor.

@@

We show how a balanced use of technology may help in the task, with a process that is necessarily multidisciplinary, joining skills coming from the diverse domains: sociology, economics, humanities, and engineering. Taking into account all such aspects is a challenging task, even when applied to a limited use case. Reporting about the design process is also a complex task, as witnessed by this paper. However, the result is rewarding and we demonstrates how the multidisciplinary interplay returns ad-hoc tools targeting sustainability and efficacy.

In the case of \emph{Casoli}, the touristic trends and the local vocation portraits a small destination for visitors practicing heterogeneous touristic circuits in individual and combined ways. Cultural sustainability mandates the involvement of local stakeholders in planning new heritage itineraries and supporting initiatives, including signage.

@@

%\nociteWeb{*}
\bibliographystyleWeb{abbrv}
\bibliographyWeb{biblio/sitography.bib}

% If authors have biography, please use the format below
\section*{Short Biography of Authors}
\bio
{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/mariagrazia.png}}}
{\textbf{Maria Grazia Deri} PhD in Migration and Tourism of the roots at the Department of Political Science of the University of Pisa. Researcher on the Geography of Tourism and Human Geography, as well as Destination and Destination Marketing, with interests in cultural and sustainable tourism. Researcher in the framework of European projects about tourism and member of the European Romea Strata Association (AERS).}

\bio
{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/letizia.png}}}
{\textbf{Letizia CHiti} Research fellow at the University of Pisa, Department of Civilization and Forms of Knowledge. She graduated in Archaeology at the University of Pisa and obtained a specialization diploma at SISBA with an essay on the archaeological map of the municipality of Bagni di Lucca (LU). He participated in projects and surface reconnaissance in mountain areas in Liguria and Tuscany, specializing in non-invasive analysis methods (GIS and study of elevations) with a diachronic approach.}

\bio
{\raisebox{-0.35cm}{\includegraphics[width=3.5cm,height=5.3cm,clip,keepaspectratio]{Definitions/augusto.png}}}
{\textbf{Augusto Ciuffoletti} Former teacher and researcher at the University of Pisa, retired on September 2023, with recent achievements on the Internet of Things and distributed control, with an interest on low-power devices. Previously active on Grid and, later, Cloud computing in collaboration with INFN. Until 2023, teaching activity at the Degree Course in Computer Humanities of the University of Pisa with two courses respectively on Network Protocols and the Angular 2+ framework.}
% For the MDPI journals use author-date citation, please follow the formatting guidelines on http://www.mdpi.com/authors/references
% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p. 328; \citeyear{ref-journal-3b}, p.475). This produces: Wong (1999, p. 328; 2000, p. 475)
